---
title: "sem_01_full"
author: "Maria_P"
date: "2026-01-05"
output: html_document
---

# Basic operations

```{r}
2 + 3
```


```{r}
sin(3.14)
```

```{r}
a <- 5
b = 5

a == b
```

```{r}
a != b + a
```

```{r}
a %in% c(1,2,3,4,5)
```

```{r}
a %in% seq(b+1, b+5)
```

```{r}
TRUE & TRUE
TRUE & FALSE
TRUE | TRUE
TRUE | FALSE
FALSE | FALSE
FALSE & FALSE
```
# Data types 

```{r}
class(a)

d <-TRUE
class(d)

f <- 'i <3 programming'
class(f)
```


# Data structures
## 1// Vectors
```{r}
vector <- c(1,2,3,'qwerty'); vector
class(vector)
```

### indexation

NB! 1 (not 0) is the 1st element!
```{r}
vector[1]
```


```{r}
vector[-1]

vector[c(1,4)]

# vector + 4 # ERROR

vector_2 <- c(1,2,3); vector_2

vector_2 + 4
```

## 2// Lists

```{r}
list_1 <- list(sample(1:13, 5)); list_1
class(list_1)

```

### indexation

```{r}
list_1[1]
list_1[[1]][3]
```

## 3// Matrix
```{r}

mtx <- matrix(sample(1:1000, 50, replace=TRUE), nrow=10, ncol=5); mtx

```

### indexation

```{r}
mtx[1,2]
mtx[1:5]

mtx[6:9, 3:4]

mtx[c(1,5,10)]

mtx[,c(3,5)]
```


## 4// DataFrame
```{r}
df <- data.frame(name=c("Маша", "Homer"),
                 age=c(25, 36),
                 city=c("Москва", "Springfield")
                 ); df
```

```{r}
dim(df)
nrow(df); ncol(df)
```

### indexation

```{r}
df[1,1]; df[1,2]; df[1,3]
```

```{r}
df$age
```


```{r}
View(df)
```


```{r}
str(df)
```


```{r}
df_2 <- data.frame(A=sample(1:10e2, size=99, replace=T)/33,
                   B=c(1,2,3),
                   C=sample(0:1, size=99, replace=T),
                   D=-5); df_2

head(df_2)
tail(df_2)

str(df_2)

```
change data type:

```{r}
df_2$C <- as.logical(df_2$C)

df_2$B <- as.factor(df_2$B)

str(df_2)
```

add new variable

```{r}
df_2$E <- round(df_2$A)
View(df_2)
```

filter data based on value

```{r}
df_2[df_2$A>10,]
df_2[(df_2$A>10) & (df_2$C) & ((df_2$E<20) | (df_2$E>=30)),]
```

## 5// Tibble

```{r}
data <- tibble::tibble(age = sample(18:60, size = 30),
               group = ifelse(age > 40, 'old', 'young'),
               group_factor = factor(group) # factor levels are defined by the alphabetic order
               ); head(data)

```


```{r}
str(data)
```

custom define factor levels

```{r}
data <- tibble::tibble(age = sample(18:60, size = 30),
                       group = ifelse(age > 40, 'old', 'young'),
                       group_factor = factor(group, levels=c('young', 'old'))
                       ); str(data)
```

 custom define factor levels and add a new level

```{r}
data <- tibble::tibble(age = sample(18:60, size = 30),
                       group = ifelse(age > 40, 'old', 'young'),
                       group_factor = factor(group, levels=c('young', 'middle', 'old'))
                       ); str(data)
```

# Tidyverse and tidy data

```{r}
library(tidyverse)
```

```{r}
data <- us_rent_income

View(data)
```

### select()
```{r}
select(data, NAME)
select(data, c(NAME, variable, estimate))
select(data, -GEOID)
```

using pipe
```{r}
data %>% select(-GEOID) -> data_selected
head(data_selected)
```

### sapply / lapply
```{r}
data_selected[1:10,] %>% select(estimate, moe) %>% sapply(mean)

data_selected[1:10,] %>% select(estimate, moe) %>% lapply(mean)

```

### Check for NAs
```{r}
is.na(data_selected)
sum(is.na(data_selected))

sapply(lapply(data_selected, is.na), sum)
# equivalent to:
data_selected %>% lapply(is.na) %>% sapply(sum)

```

How to deal with NAs?

```{r}
data_selected[is.na(data_selected$estimate),]

data_selected[is.na(data_selected$estimate), c("estimate", "moe")] <- 0
data_selected %>% lapply(is.na) %>% sapply(sum)

data_selected %>% fill(c(estimate, moe), .direction = "up") %>% lapply(is.na) %>% sapply(sum)
```

### filter()

```{r}
data_selected %>% filter(NAME=='Alaska')

data_selected %>% filter(NAME %in% c('Alaska', 'Nevada', 'Texas')) -> data_filtered

```

### mutate()
```{r}
data_filtered %>% mutate(
  first_letter = startsWith(NAME, 'N')
  )

```

### *_join()

```{r}
data_filtered %>% select(-moe) -> data_left
data_filtered[1:3,] %>% select(-estimate) -> data_right
```


```{r}
left_join(data_left, data_right)
```



```{r}
right_join(data_left, data_right)
```

```{r}
full_join(data_left, data_right)
```


```{r}
inner_join(data_left, data_right)
```


```{r}
anti_join(data_left, data_right)
```


```{r}
data_left$id <- 1:6
data_right$id <- 4:6
left_join(data_left, data_right, by='id')
```

### groupping and aggregation

```{r}
data_filtered %>% group_by(NAME)
```


```{r}
data_filtered %>% group_by(NAME) %>% summarize(mean_moe=mean(moe),
                                               median_moe = median(moe),
                                               sd_moe=sd(moe),
                                               min_moe=min(moe),
                                               max_moe=max(moe))
```


```{r}
data_filtered %>% group_by(variable) %>% summarize(mean_moe=mean(moe),
                                               median_moe = median(moe),
                                               sd_moe=sd(moe),
                                               min_moe=min(moe),
                                               max_moe=max(moe)) -> data_groupped
```

### Long vs. Wide format

```{r}
data_selected %>% select(-moe) %>% pivot_wider(names_from = variable, values_from = c(estimate)) -> data_wide
data_wide
```


```{r}
data_wide %>% pivot_longer(cols=2:3, names_to='variable', values_to='number')

```


```{r}
# equivalent to
data_wide %>% pivot_longer(cols=c('income', 'rent'), names_to='variable', values_to='number')
```


```{r}
# equivalent to
data_wide %>% pivot_longer(cols=colnames(data_wide)[2:3], names_to='variable', values_to='number')
```



